---
title: "chapter_3_problems"
author: "Dustin Duncan"
date: "2024-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rethinking)
rm(list = ls())
```

Using this code to answer our questions 
```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)

prior <- rep(1, 1000)

likelihood <- dbinom(6, size = 9, prob = p_grid)

posterior <- likelihood * prior

posterior <- posterior / sum(posterior)

set.seed(100)

samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)
```

### 3E1 

How much posterior probability lies below p = 0.2?
```{r}
sum(samples < 0.2)/1e4
```

About .08% of our posterior probability lies below p = 0.2

### 3E2

How much posterior probability lies above 0.8? 
```{r}
sum(samples > 0.8)/1e4
```

About 12% 

### 3E3

How much posterior probability lies betwen 0.2 and 0.8? 
```{r}
sum(samples > 0.2 & samples < 0.8) / 1e4
```

About 87% 

### 3E4

20% of the posterior probability lies below which value of p? 
```{r}
quantile(samples, 0.2)
```

20% of the posterior probability lies below p = 0.51

### 3E5

20% of the posterior probability lies above which value of p?
```{r}
quantile(samples, 1 - 0.2)
```

20% of posterior probability lies above p = 0.76

### 3E6

Which values of p contain the narrowest interval equal to 66% of the posterior probability? 
```{r}
HPDI(samples, prob = 0.66)
```

p = 0.51 and p = 0.77

### 3E7

Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval? 
```{r}
PI(samples, prob = 0.66)
```

0.50 and 0.77


### 3M1

Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution using grid approximation. Use the same flat prior as before
```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)

prior <- rep(1, 1000)

likelihood <- dbinom(8, size = 15, prob = p_grid)

posterior <- likelihood * prior

posterior <- posterior / sum(posterior)
```


### 3M2 

Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p
```{r}
# Draw 10,000 samples
set.seed(100)

samples <- sample(p_grid, prob = posterior, size = 10000, replace = TRUE)
```

```{r}
# 90% HPDI

HPDI(samples, prob = 0.9)
```

90% HPDI lies between p = 0.33 and p = 0.72

### 3M3
Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses? 
```{r}
w <- rbinom(1e4, size = 15, prob = samples)
simplehist(w)
```

```{r}
mean(w == 8)
```

The probability of observing 8 water in 15 tosses is about 15% 

### 3M4

Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses
```{r}
w2 <- rbinom(1e4, size = 9, prob = samples)
mean(w2 == 6)
```

### 3M5

Start over at 3M1, but now use a prior that is zero below p = 0.5 and a constant above p = 0.5. This corresponds to prior information that a majority of the earths surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value p = 0.7
```{r}
p_grid <- seq(from = 0, to = 1, length.out = 1000)

prior <- ifelse(p_grid < 0.5, 0, 1)

likelihood <- dbinom(8, size = 15, prob = p_grid)

posterior <- likelihood * prior

posterior <- posterior / sum(posterior)

samples2 <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)

plot(x = p_grid, y = posterior, type = "l")
```

Calculating 90% HPDI 
```{r}
HPDI(samples2, prob = 0.90)
```

Constructing a posterior predictive check and plotting off samples. Then finding the probability of 8 water in 15 tosses.
```{r}
w <- rbinom(10000, size = 15, prob = samples2)
simplehist(w)
```
Probability of getting 8 water in 15 tosses. 
```{r}
mean(w == 8)
```


```{r}
w <- rbinom(1e4, size = 9, prob = samples2)
mean(w == 6)
```

The new prior increased the posterior probability of the true parameter being p = 0.7

### 3M6
Suppose you want to estimate the Earth's proportion of water very precisely. Specifically, you want the 99% percentile interval of the posterior distribution of p to be only 0.05 wide. This means that distance between the upper and lower bound of the interval should be 0.05. How many times will you have to toss the globe to do this? 

You will have to toss the globe many times. 

### Hard
The hard problems here all use the data below. These data indicate the gender (male=1, female=0) of officially reported first and second born children in 100 two-child families. 
```{r}
birth1 <- c(1,0,0,0,1,1,0,1,0,1,0,0,1,1,0,1,1,0,0,0,1,0,0,0,1,0,


0,0,0,1,1,1,0,1,0,1,1,1,0,1,0,1,1,0,1,0,0,1,1,0,1,0,0,0,0,0,0,0,


1,1,0,1,0,0,1,0,0,0,1,0,0,1,1,1,1,0,1,0,1,1,1,1,1,0,0,1,0,1,1,0,


1,0,1,1,1,0,1,1,1,1)

birth2 <- c(0,1,0,1,0,1,1,1,0,0,1,1,1,1,1,0,0,1,1,1,0,0,1,1,1,0,


1,1,1,0,1,1,1,0,1,0,0,1,1,1,1,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1,1,1,


1,1,1,0,1,1,0,1,1,0,1,1,1,0,0,0,0,0,0,1,0,0,0,1,1,0,0,1,0,0,1,1,


0,0,0,1,1,1,0,0,0,0)

```

So for example, the first family in the data reported a boy (1), and then a girl (0). The second family reported a girl (0) and then a boy (1). The third family reported two girls. You can load these two vectors into R's memory by typing:

```{r}
data(homeworkch3)
```

Use these vectors as data. So for example to compute the total number of boys born across all of these births, you could use. 
```{r}
sum(birth1) + sum(birth2)
```


### 3H1
Using the grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior probability? 
```{r}
allbirths <- c(birth1, birth2)

p_grid <- seq(from = 0, to = 1, length.out = 1000)

prior <- rep(1, 1000)

likelihood <- dbinom(sum(allbirths), size = length(allbirths), prob = p_grid)

posterior <- likelihood * prior 

posterior <- posterior / sum(posterior)

plot(x = p_grid, y = posterior, type = "l")
```

```{r}
p_grid[which.max(posterior)]
```

When p = 0.55, it maximizes the posterior probability.

### 3H2
Draw 10,000 random samples from the posterior distribution. use these samples to estimate the 50%, 89%, and 97% highest posterior density intervals. 
```{r}
samples <- sample(p_grid, prob = posterior, size = 1e4, replace = TRUE)

# 50% HPDI
HPDI(samples, prob = 0.5)

# 87% HPDI
HPDI(samples, prob = 0.87)

# 97% HPDI
HPDI(samples, prob = 0.97)
```

### 3H3
Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distribution of predicted numbers of boys to the actual count in the data (111 out of 200 births). There are many good ways to visualize the simulation, but the dens function is probably the easiest way in this case. Does it look like the model fits the data well? That is, does the distribution of predictions include the actual observation as a central, likely outcome?

simulating 10,000 replicates of 200 births. 
```{r}
counts <- rbinom(1e4, size = 200, prob = samples)
dens(counts)
abline(v = sum(allbirths), col = "red")
```

Yes this does seem to fit the data well, as the modal value is around 111, the actual value of boy births. 
### 3H4
Now compare 10,000 counts of boys from 100 simulated first borns only to the number of boys in the first births, birth1. How does the model look in this light?

```{r}
likelihood <- dbinom(sum(birth1), size = length(birth1), prob = p_grid)

posterior <- likelihood * prior

posterior <- posterior / sum(posterior)

samples <- sample(p_grid, prob = posterior, size = 100, replace = TRUE)

counts1 <- rbinom(1e4, size = 100, prob = samples)
dens(counts1)
abline(v = sum(birth1), col = "green")
```

The observed value of 51 boys in 100 first borns is consistent with the model 

### 3H5
The model assumes that sex of first and second births are independent. To check this assumption, focus now on second births that followed female first borns. Compare 10,000 simulated counts of boys to only those second births that followed girls. To do this correctly, you need to count the number of first borns who were girls and simulate that many births 10,000 times. Compare the counts of boys in your simulations to the actual observed count of boys following girls. How does the model look in this light? Any guesses as to what is going on in these data? 

```{r}
nfirstgirls <- sum(birth1 == 0)

counts2 <- rbinom(1e4, size = nfirstgirls, prob = samples)

dens(counts2)

ngirlthenboy <- sum(birth2[birth1 == 0])

abline(v = sum(ngirlthenboy), col = "blue")
```

The fact that the observed value is so far above the majority of the data suggests that this model is a poor fit to the data. This indicates that the sex of first and second born children are not independent. This may suggest that males are more likely to follow female first-borns.
